---
title: "Mapping FBI Crime Data Reporting Across the U.S."
subtitle: "Proposal"
author: 
  - name: "Ashton Norman"
    affiliations:
      - name: "College of Engineering, University of Arizona"
description: "This project will explore patterns in FBI crime reporting participation using the TidyTuesday dataset on agency participation in the National Incident-Based Reporting System (NIBRS)."
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  echo: false
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
#| warning: false

#Install packman once in Console:
#install.packages("pacman")

pacman::p_load(
  tidyverse,
  tidytuesdayR, #used to load the dataset
  scales, #used to format percentages 
  gt, #table formatting https://gt.rstudio.com/ - articles followed for usage
  usmap) #US map package, reference manual and vignettes followed for usage https://cran.r-project.org/web/packages/usmap/index.html 
```

## Dataset

```{r}
#| label: Dataset Load & Preview
#| message: false
#Load agency reporting data
agencies <- tidytuesdayR::tt_load(2025, week=7)$agencies

#Alternative - load from .csv located in the data folder
#agencies <- read_csv("data/agencies.csv") 

head(agencies)

#Setting up variables from the data
state <- agencies$state
county <- agencies$county
latitude <- agencies$latitude
longitude <- agencies$longitude
agency_type <- agencies$agency_type
agency_ori <- agencies$ori
report_status <- agencies$is_nibrs
start_date <- agencies$nibrs_start_date

```

This dataset was sourced from the [FBI's Crime Data API](https://cde.ucr.cjis.gov/LATEST/webapp/#/pages/docApi) and curated by [Ford Johnson](https://github.com/bradfordjohnson) for the [TidyTuesday project on February 18, 2025](https://github.com/rfordatascience/tidytuesday/tree/main/data/2025/2025-02-18). The data includes law enforcement participation in the National Incident-Based Reporting System (NIBRS), agency names, type (county/city), geographic location, and reporting start date.

#### Dataset Overview

```{r}

#| label: dataset-overviewl
#| message: false

library(glue)
# Summarize basic information about the dataset

percent_reporting <- percent(sum(report_status=="TRUE", na.rm=TRUE)/nrow(agencies))

agencies_vars <- c("County", "State", "Agency Type", "Agency (ORI)", "Reporting", "Start Date")
agencies_class <- c(class(county), class(state), class(agency_type), class(report_status), class(start_date))

notes <- c(
  glue("Number of counties included: {length(unique(county))}"), 
  glue("Number of states included: {length(unique(state))}"), 
  paste(na.omit(unique(agency_type)), collapse=', '), 
  glue("Number of agencies included: {length(unique(agency_ori))}"),
  glue("Percentage of agencies reporting: {percent_reporting}"),
  paste("Date Range: ", min(start_date, na.rm=TRUE), "to", max(start_date, na.rm=TRUE)))

agencies_summary <- data.frame(agencies_vars, notes)

colnames(agencies_summary) <- c("Variables", "Notes")

gt(agencies_summary) |>
  tab_header(
    title = "Dataset Overview") |>
  tab_style(
    styl=cell_text(weight="bold"),
    locations=cells_column_labels()
  )
view(agencies)
```

```{r fig.height=10}

#| label: Participation by state
#| message: false

state_agencies_all <- agencies |>
 count(state)  

state_agencies_report <- agencies |>
 filter(report_status=="TRUE") |>
  count(state)

state_agencies <- merge(state_agencies_all, state_agencies_report, by="state")

state_agencies$state_agencies_percent <- percent(state_agencies$n.y/state_agencies$n.x, accuracy=1)

colnames(state_agencies) <- c("State", "Total Agencies", "Agencies Reporting", "Percent Participating")

ggplot(state_agencies, aes(x=reorder(State, `Total Agencies`)))+
  geom_col(aes(y=`Total Agencies`, fill="Total Agencies"))+
  geom_col(aes(y=`Agencies Reporting`, fill="Agencies Reporting"))+
  geom_text(aes(
    y=`Total Agencies`+100,
    label=`Percent Participating`),
    size=3,
    color="gray30")+
  coord_flip()+
   scale_fill_manual(
     values = c("Total Agencies" = "lightcyan3","Agencies Reporting"="darkcyan"))+
  labs(
    x=NULL,
    y="Number of Agencies",
    title="Agency NIBRS Participation by State",
    caption="Bar chart showing the number of law enforcement agencies in each state and number participating in the National Incident-Based Reporting System (NIBRS). Labels indicate the percentage of agencies in each state currently reporting to NIBRS",
    fill=NULL
  )+
  theme_minimal()

```

I selected this dataset because it includes geographic variables such as `longitude`, `latitude`, `state`, and `county` that can be used for location-specific and state-level map visualizations. Participating agencies also have a `start_date`, which will allow for visualizations of timelines and pattern changes. My work involves research on infectious diseases, so improving my skills in these areas would be useful for communicating information such as outbreak data trends.

## Questions

1.  Which U.S. states and regions have lower levels of NIBRS adoption by law enforcement agencies, and are there patterns by geography or agency type?
    1.  Hypothesis: States with a mix of large cities and rural areas will have lower NIBRS participation due to variation in agency structure and likelihood of having a large number of total agencies.
2.  How has NIBRS participation expanded geographically from 1985 to 2024?
    1.  Hypothesis: NIBRS adoption has increased more rapidly in the last decade due to advancements in technology and broader adoption of centralized and streamlined data collection and reporting systems. I expect states with smaller populations to have higher adoption rates earlier, while more populous states will lag behind.

## Analysis plan

-   Question 1: Current participation levels have been explored above at the state level and graphed on the bar chart titled *Agency Participation by State,* which compares total and number of participating agencies within states and includes the percentage reporting to NIBRS as labels. I plan to build on this to further explore patterns and trends by agency type and region (Northeast, Midwest, South, and West).
    -   Quantitative Methods: Calculate proportion of participating agencies (`report_status==TRUE`) within each `region` and by `agency_type`, if time permits, significance tests may be explored.
    -   Visual Methods: Bar or dot plots for side-by-side comparisons - other options for visual comparison may be explored as well.
    -   Variables:
        -   `state`: to summarize participation by state and region
        -   `county`: to potentially be used for a more detailed look at specific areas within states or across certain regions
        -   `agency_type`: for comparing participation across agency types
        -   `report_status`: variable for determining participation status
        -   `region`: states grouped to 4 US Census regions based on definitions referenced from the [CDC](https://www.cdc.gov/nchs/hus/sources-definitions/geographic-region.htm) website. Full list and map of states by region included below:

```{r}
#| label: US Regions List
#| message: false


region_states <- tibble::tibble(
  Region = c("Northeast", "Midwest", "South", "West"),
  States = c(
    "Connecticut, Maine, Massachusetts, New Hampshire, New Jersey, New York, Pennsylvania, Rhode Island, Vermont",
    "Illinois, Indiana, Iowa, Kansas, Michigan, Minnesota, Missouri, Nebraska, North Dakota, Ohio, South Dakota, Wisconsin", 
    "Alabama, Arkansas, Delaware, Florida, Georgia, Kentucky, Louisiana, Maryland, Mississippi, North Carolina, Oklahoma, South Carolina, Tennessee, Texas, Virginia, West Virginia",
    "Alaska, Arizona, California, Colorado, Hawaii, Idaho, Montana, Nevada, New Mexico, Oregon, Utah, Washington, Wyoming"))

region_states |>
  gt() |>
  tab_header(title = "U.S. Regions") |>
  tab_source_note(
    source_note = md("*Regions defined by the CDC based on U.S. Census Bureau groupings,Source: https://www.cdc.gov/nchs/hus/sources-definitions/geographic-region.htm)*")) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels())

  
```

```{r}
#| label: US Regions Map
#| message: false

#add regions column to agencies data frame
agencies <- agencies |>
  mutate(region=case_when(
    state %in% c("Connecticut", "Maine", "Massachusetts", "New Hampshire", "New Jersey", "New York", "Pennsylvania", "Rhode Island", "Vermont") ~ "Northeast",
    state %in% c("Illinois", "Indiana", "Iowa", "Kansas", "Michigan", "Minnesota", "Missouri", "Nebraska", "North Dakota", "Ohio", "South Dakota", "Wisconsin") ~ "Midwest",
    state %in% c("Alabama", "Arkansas", "Delaware", "Florida", "Georgia", "Kentucky", "Louisiana", "Maryland", "Mississippi", "North Carolina", "Oklahoma", "South Carolina", "Tennessee", "Texas", "Virginia", "West Virginia") ~ "South",
    state %in% c("Alaska", "Arizona", "California", "Colorado", "Hawaii", "Idaho", "Montana", "Nevada", "New Mexico", "Oregon", "Utah", "Washington", "Wyoming") ~ "West"))

agencies |>
  distinct(state, region) |>
  plot_usmap(
  data=_,
    values = "region",
    exclude=c("District of Columbia"), #DC not in agencies so it needs to be explicitly excluded when using usmap
    color = "white")+ 
  scale_fill_manual(
    name = "Region", 
    values=c(
      "Northeast"="steelblue3",
      "Midwest"="mediumorchid",
      "South"="deeppink2",
      "West"="coral")) + 
  labs(title = "US Regions")+
  theme(legend.position = "right")
```

-   Question 2: Similar to Question 1, I will build off of the current participation calculations to analyze participation expansion over time. While the `usmap` package was utilized to create the map above, but I do not believe it has coordinate plotting capabilities. Alternatives will be explored but I may change to county plotting if needed for time restraints.
    -   Quantitative Methods: Rates of adoption will be calculated for each year and/or time period - potentially as trend lines for general direction and rates of adoption or growth curves for cumulative change.
    -   Visual Methods: Faceted maps to show changes in participation over time covering 1985-2024 (likely groups of either 5 or 10 years - e.g., 1985-1994, 1995-2004, 2005-2014, 2015-2024) - may also explore animation if possible. Depending on the final quantitative methods, other visualizations such as line charts, bar plots, or cumulative growth curves may be included to help represent results over time.
    -   Variables:
        -   `start_date`: will provide years when agencies began reporting to NIBRS

        -   `latitude` and `longitude`: will be used for mapping agency locations

            -   `county` will be used as an alternative if using `usmap` package

        -   `report_status`: variable for determining participation status

## Plan of Attack

```{r}
#| label: Participation by state
#| message: false

timeline <- tibble::tibble(
  `Task Name`=c("Proposal draft", 
                "Resolve peer/instructor review", 
                "Data wrangling", 
                "Q1 Analysis calculations", 
                "Q1 Complete visualizations", 
                "Q2 Analysis calculations", 
                "Q2 Complete visualizations", 
                "Finalize project and presentation", 
                "Submit final report"),
  `Status`=c("Complete",
             "Complete",
             "In progress",
             "In progress",
             "In progress",
             "To be completed",
             "To be completed",
             "To be completed",
             "To be completed"
             ),
  `Due`=c("06/10/25",
          "06/18/25",
          "06/21/25",
          "06/22/25",
          "06/24/25",
          "06/28/25",
          "06/29/25",
          "07/02/25",
          "07/02/25"),
  `Priority`=c(
    "High",
    "High",
    "Moderate",
    "Moderate",
    "Moderate",
    "Moderate",
    "Moderate",
    "High",
    "High"),
  `Summary`=c(
    "Draft proposal with questions, analysis plan, and exploratory analysis/visualization",
    "Address feedback received during peer/instructor review",
    "Prepare data for analysis, finalize required variables",
    "Calculate participation by area and agency type",
    "Create plots and summaries by agency type and region",
    "Calculate rates of adoption over time",
    "Create maps and time-based plots",
    "Complete and polish results summaries and interpretation; finish and record presentation",
    "Submit final report, recorded presentation, and website"
  ))

timeline |>
  gt() |>
  tab_header(title = "Plan of Attack Timeline") |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels())

```
